---
output: 
  github_document:
    toc: true
    toc_depth: 4
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- knitr::knit("README.Rmd") -->
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dev.args = list(png = list(type = "cairo"))
)
```

# HexTemplatesFCS

<!-- badges: start -->
<!-- badges: end -->

The goal of HexTemplatesFCS is to quickly and easily generate data structures that make it easy to get statistical output starting from raw fcs-files and a metadata table. The grid approach is especially useful if there aren't any populations clearly visible in the files.


## Installation

You can install the development version of HexTemplatesFCS from GitHub with:

``` r
library(devtools)
install_github("AG-ESSER/HexTemplatesFCS")
```

## Quick Start

### Reading Files and Metadata

First we read the fcs-files and the metadata table.
```{r quickstart}
library(flowCore)

setwd("C:/Users/student.esser/FACSDATA/FCS_skin_microbiome")  #set working directory
fcs <- read.flowSet(pattern = ".fcs")                         #read all .fcs-files in directory
metadata <- read.table("metadata.csv", sep = ";", header = T) #read metadata table
head(metadata)
```


### Creating the HexTemplate
Check the available channels:
```{r}
colnames(fcs)
```

Next we create the HexTemplate. Fine tuning of the *HexTemplate* function is explained in the "HexTemplate"-chapter.
```{r}
library(HexTemplatesFCS)

hexT <- HexTemplate(flowset = fcs, xChannel = "SSC-H", yChannel = "FITC-H", metadata = metadata)
```

### Creating the Distance Matrix

We can create a distance matrix directly from the HexTemplate using the *weightedBray* function. It automatically adjusts for spatial dependencies in the hexagon grid by assigning weights to every combination of hexagons based on their distance from another. Further Details and how to create a distance matrix without adjustments are explained in the "Weights and Distance Matrices"-chapter.
```{r}
distM <- weightedBray(hexT)
```

### Analysis of Variance using ADONIS

The function *adonis* from the *vegan* package makes it possible to perform an analysis of variance on a distance matrix to test for significant effects in treatment groups. We can use the *meta.data* function to access the metadata slot of the HexTemplate.

```{r}
library(vegan)

adonis(distM ~ areal + sex, data = meta.data(hexT))
```

### nmds-plot
We can see that there is a significant difference between the samples of the areal metadata group. To visualize the difference between all of the samples we can use a nmds-plot. For this we can use the *metaMDS* function that is also included in the *vegan* package.

```{r, message = F}
nmd <- metaMDS(distM, trace = F)

ordiplot(nmd, display = "sites")
ordispider(nmd, groups = meta.data(hexT)[,"areal"])
ordiellipse(nmd, groups = meta.data(hexT)[,"areal"], label = T, draw = "polygon")
```

### Visualizing regional Differences with t-scores
Now we might want to know if the significant effect stems from a certain region within the SSC-H-FITC-H area. For this we calculate t-scores based on Welch's t-statistic independently for every hexagon and for every combination of levels in a given metadata group and use them as a measurement of difference. 

The *tscores* function does this automatically. It returns a data.frame of t-scores with the columns representing every possible unique combination of levels in the given metadata group. In this case we have two levels (Areal1 and Areal2) so the output will be a data.frame with a single column.

```{r}
ts <- tscores(hexT, "areal")
```

We can visualize the differences between two levels of a metadata group using the *plot_tscores* function. We don't have to worry about indexing ts because there is only one possible combination of levels.
```{r}
plot_tscores(hexT, ts, color = NA)
```
In this case a blue colored hexagon indicates a decrease in event percentage in this region in the samples of the Areal2 level compared to Areal1. The big blue region might be something we want to look at more closely and possibly perform a subsequent FACS analysis where this area could be sorted and further evaluated. How to automatically detect gates based on the t-score and how to integrate them in FlowJo-workspaces is explained in the "t-Score Gate Detection"-chapter.

## HexTemplate

There are some adjustments we can do to the HexTemplates function. We can change the resolution of the hexagon grid by using the **xbins** argument that is directly passed to the *hexbin* function from the *hexbin* package that HexTemplate uses internally to create the grid. Default is 20 (meaning 21 hexagons on the x-axis). This number should not be lowered as the *adonis* and *metaMDS* functions are not stable for lower numbers of hexagons.

Furthermore we can give a function to transform the data to the **trans_fun** argument. If no function is given the data will be log10 transformed. If the data should not be transformed at all then *trans_fun = return* will do the trick.

The following creates the same HexTemplate that was created in the "Quick Start"-chapter.

```{r, eval = FALSE}
hexT <- HexTemplate(flowset = fcs,
                    xChannel = "SSC-H",
                    yChannel = "FITC-H",
                    metadata = metadata,
                    xbins = 20,
                    trans_fun = log10)
```

## Plotting
The S3 *plot* method for the HexTemplate class and the *plot_tscores* function return *ggplot* objects and are therefore highly customizable. They use geom_sf internally to display the hexagons. Additional arguments passed to geom_sf can be added to the function call and additional layers can be added dynamically as to any *ggplot* object. 

We can plot the empty grid:
```{r}
plot(hexT)
```

Or fill it with the event percentages of sample no. 5
```{r}
plot(hexT, sample = 5)
```

If additional layers should be added to the plot we first have to load the *ggplot2* library.

Plot the id of every hexagon
```{r}
library(ggplot2)

plot(hexT) + 
  geom_sf_text(mapping = aes(label = 1:hexT@nHex), size = 2.5)
```

Plot the number of events in every hexagon for sample no. 5 and hide the legend
```{r}
plot(hexT, 5, color = NA) + 
  geom_sf_text(mapping = aes(label = hexT@counts[[5]]), size = 2) +
  theme(legend.position = "none")
```

## Weights and Distance Matrices

### Distance Matrix without adjustments for spatial dependencies

First we need to turn the absolute event numbers contained in the HexTemplate into frequencies. We can do that using the *frequencies* function. Next we can create a distance matrix using the *vegdist* function from the *vegan* package in conjunction with the frequency table. 

```{r, eval = FALSE}
freq <- frequencies(hexT)

library(vegan)
distM_vegan <- vegdist(freq, method = "bray")
```


### The weightedBray function

The default way to create a distance matrix from a HexTemplate is the *weightedBray* function which expands on the Bray-distance by adding in a weight matrix that contains a weight for every combination of hexagons. We can give the function a custom weight matrix but the default weight matrix is calculated with exp(-1\*gamma\*euclDist) where gamma is a factor controlling the speed of the decline and euclDist is a matrix containing the euclidean distances between the hexagon's centers. We can control the gamma factor directly with the **gamma** argument of the *weightedBray* function. Default is 8. Lower values increase the region around any hexagon that is taken into account when calculating the distance matrix. Higher values make the function behave closer to the default Bray-distance.

```{r, eval = FALSE}
distM8 <- weightedBray(hexT, gamma = 8)

distM2 <- weightedBray(hexT, gamma = 2) #bigger region of spatial dependecy

```

### Custom Weight Matrices

If the default exponential approach is not what we want we can use the *weightMatrix* function and give it to the **w** argument of the *weightedBray* function. For the *weightMatrix* function we can choose between an exponential approach or a discrete approach by calling the function with either "exp" or "disc" as the **method** argument. If no custom weight matrix is given then *weightedBray* uses the exponential approach of the *weightMatrix* function by default.

The following creates the same distance Matrix as in the "Quick Start"-chapter:
```{r, eval = FALSE}
wM_exp <- weightMatrix(hexT, method = "exp", gamma = 8)
distM <- weightedBray(hexT, w = wM_exp)
```

The discrete approach gives us more control over the final matrix by letting us give a weight for every ranked distance between hexagons to the **val** argument of the *weightMatrix* function. However this approach depends on the resolution of the hexagon grid.

```{r, eval = FALSE}
wM_disc <- weightMatrix(hexT, method = "disc", val = c(.5,.25,.25))
```

### Visualizing Weight Matrices

We can visualize the different weight matrices by plotting the empty hexagon grid and giving the a column of a weight matrix to the fill aesthetic of the *plot* function.

In this case we plot the weights of every hexagon relating to hexagon no. 256 as contained in the default exponential weight matrix.
```{r}
wM <- weightMatrix(hexT, method = "exp", gamma = 8)

plot(hexT, mapping = aes(fill = wM[,256])) + 
  geom_sf_text(aes(label = 1:hexT@nHex), size = 2)
```

## t-Score Gate Detection

### detectGates

We can use the *detectGates* function to try and approximate meaningful gates based on the t-scores using a floodfill approach.
```{r}
gates <- detectGates(hexT, ts)
```

We can overlay the gates over the t-score plot by turning the gates into simple features polygons.
```{r}
library(sf)

psf <- st_polygon(gates)
```

Plot the gates and t-scores.
```{r}
plot_tscores(hexT, ts, color = NA) + 
  geom_sf(data = psf, fill = NA)
```

There is a lot of fine tuning that we can do for the detectGates function. The function is a simple visual aid and the starting hexagons are randomly seeded so we can feel free to try multiple times and with different arguments until we have a seed and configuration that we like. We can check the documentation of the detectGates function by typing '?detectGates' into the R-console.

A more sensible configuration
```{r}
gates <- detectGates(hexT, ts, conc = 5, thresh = 3, nM = 3, tSeed = 248)
```

Turn into simple features object and plot
```{r}
psf <- st_polygon(gates)

plot_tscores(hexT, ts, color = NA) + 
  geom_sf(data = psf, fill = NA)

```

### Create FlowJo workspaces with the t-score based gates

For this we need the *flowWorkspace* package, the *CytoML* package in addition to the *flowCore* package as well as an installation of docker which runs in the background. Install docker and then follow the steps here to get the *gatingset_to_flowjo* function working: [https://hub.docker.com/r/rglab/gs-to-flowjo](https://hub.docker.com/r/rglab/gs-to-flowjo).

First we create a function for transforming our data. This needs to be the inverse of the original transformation function that was given to the HexTemplate function in the beginning. Here it was log10 so our inverse function looks like this:

```{r, eval = FALSE}
inv <- function(x) {
  10^x
}
```

Next we turn our flowSet into a GatingSet and apply our gates to it. Finally we turn it into a FlowJo workspace by running the *gatingset_to_flowjo* function while docker is running in the background.

```{r, eval = FALSE}
library(flowWorkspace)

gs <- GatingSet(fcs)

for (i in 1:length(gates)) {
  sqrcut <- inv(gates[[i]])
  colnames(sqrcut) <- c(hexT@xChannel, hexT@yChannel)
  pg <- polygonGate(filterId = as.character(i), .gate = sqrcut)
  gs_pop_add(gs, pg)
}

recompute(gs)

library(CytoML)
outFile <- paste("skin_microbiome", fileext = ".wsp", sep = "")
gatingset_to_flowjo(gs, outFile)
```
